Evaluating Sound Similarity Metrics for Differentiable, Iterative Sound-Matching

### 2025-08-26 revised_main.tex
Updates from Overleaf

\documentclass[lettersize,journal]{IEEEtran}
\usepackage{amsmath,amsfonts}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{array}
\usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
\usepackage{textcomp}
\usepackage{stfloats}
\usepackage{url}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{csquotes}
\usepackage{makecell}
% my added packages and commands =======================
\usepackage[numbers]{natbib} %removed cite package

@@

% \subsection{Sound Representation and Loss Functions}
% \label{sec:loss_funcs}
% A digital sound (or an audio signal) is a series of numbers~\cite{smith1991viewpoints,smith2007mathematics}. To compare two digital sounds, the two corresponding series are passed to a function that measures their similarity. Two signals can sound identical to our ears, without having any values in common~\cite{moore2012introduction}. This necessitates the use of proxy representations (or feature extractors) when comparing sounds automatically. Similarity between the target sound and the synthesizer output is then measured by some form of subtraction and summation of the proxy representations.

% In sound-matching, particularly in a Deep Learning (\gls{DL}) context~\cite{goodfellow2016deep}, the similarity function can also be called a \textit{loss} function, where the emphasis is on measurement and reduction of the distance between target and output. It is important to note that there is a close relationship between the loss function $L$ and the sound representation function $\phi$. $L$ is the result of a distance measure $d$ applied to the features extracted by $\phi$.

% \[
% L(\theta, t) = d\langle\phi(t),\phi(x)\rangle
% \]

% \noindent

% A proxy representation is the output of the function \( \phi \), which can be thought of as a feature extraction function that maps the sounds \( t \) and \( x \) to their respective representations. 
% The proportionality or distance metric $d$ has typically been the L1 distance~\cite{turian2020sorry,richard2025model}, calculated as the mean of the absolute difference between every point in the proxy representation~\cite{engel2020ddsp,vahidi2023mesostructures}:
% \[
% L(\theta, t) = \left\| \phi(t) - \phi(x) \right\|_1
% \]

% Here we discuss four methods of audio representation and the corresponding loss functions. Three of which have been used in previous works, and a novel method not previously used in iterative sound-matching.

% \subsubsection{Parameter Loss}
% A common measure of similarity in sound-matching is the distance between synthesizer parameter sets, referred to as ``P-Loss"~\cite{han2023perceptual}. Typically, for the implementation of P-Loss the parameter sets are treated as vectors in space, and L1 or L2 distance is applied. There are two major limitations to this approach: First, the target and output sound must be made by the same synthesizer; otherwise the parameter sets cannot be compared (see Section~\ref{sec:matching_types}). Second, the relationship between synthesizer parameters and the audio output is not linear~\cite{shier2020spiegelib,han2023perceptual,esling2019flow}.

% \subsubsection{Fourier Spectrograms}
% \label{sec:fourier_specs}
% Fourier-based transformations such as short-time Fourier transforms (\gls{STFT}), Mel-spectrograms, and Mel-frequency cepstral coefficients have been viewed as the de facto and state-of-the-art representation of audio~\cite{beauchamp2003error,mitchell2007evolutionary,yee2018automatic}, however, there are many issues associated with their use in sound-matching~\cite{turian2020sorry,vahidi2023mesostructures,han2023perceptual,uzrad2024diffmoog}. Fourier transformations allow for the conversion of a signal from the time-domain to the frequency domain. Audio spectrograms can be generated by segmentation of a piece of audio into overlapping windows followed by the application of Fourier transforms to each window. They are costly to compute, but provide a better temporal view of changes in frequency content~\cite{muller2007dynamic,smith2007mathematics}. There are different types of spectrograms that have a basis in Fourier transformations, but the most notable and commonly used is the STFT.

% What we call \textit{Fourier-based Spectrograms} are variations on the STFT approach. For example, Mel-Spectrograms \textit{bin} frequencies on a near-logarithmic scale to better match human perception of frequencies~\cite{muller2007dynamic}. Multi-scale spectrograms (\gls{MSS}) used in recent works are a simple weighted average of multiple spectrograms with different parameters such as window size, number of frequency bins, and hop size~\cite{engel2020ddsp,vahidi2023mesostructures}; this may provide some improvements at a higher computational cost~\cite{turian2020sorry,engel2020ddsp}.

% \subsubsection{Joint-Time Frequency Spectrum}
% Recent works have focused on the limitations of parameter and spectral loss functions in sound-matching~\cite{vahidi2023mesostructures,uzrad2024diffmoog}, seeking to create more effective general solutions for the comparison of audio. 
% Noting the aforementioned weaknesses of comparing STFT spectrograms, Vahidi \textit{et al.} proposed differentiable Joint-Time Frequency Scattering (\gls{JTFS})~\cite{anden2015joint} as an alternative to spectrogram loss in sound-matching, and showed improved performance in sound-matching with differentiable chirplet synthesizers~\cite{vahidi2023mesostructures}. JTFS is the result of the application of a 2D wavelet transformation to the time-frequency representation of a signal~\cite{anden2015joint}.

% \subsubsection{Dynamic Envelope Warping}
% Dynamic Time Warping (DTW) is a method for measuring similarity between multi-dimensional time-series~\cite{rabiner1993fundamentals,muller2007dynamic,giorgino2009computing}. Given any two time-series $X = \{x_1,x_2,...,x_m\}$ and $Y = \{y_1,y_2,...,y_n\}$, we have indices $i\in\{1...m\}$ and $j\in\{1...n\}$ defining $X$ and $Y$. When the series are \textit{warped}, these indices change to expand or contract different portions of the series. To borrow the notation given by Muller~\cite{muller2007dynamic}, warped indices are a sequence $p=(p_1,...,p_L)$, where \(p_\ell = (m_\ell, n_\ell) \in [1 : m] \times [1 : n] \text{ for } \ell \in [1 : L]\), meaning that the indices for $X$ and $Y$ are reorganized under special conditions. In classical DTW, these conditions are \textit{monotonicity}, \textit{boundary matching}, and \textit{single step-size}.

% DTW measures the distance between the time-series \textit{after} alignments, typically using Euclidean distance. The distance between a time-series and shifted versions of itself would be 0, regardless of shift amount~\cite{tavenard.blog.dtw}. Additional rules can be imposed to keep alignments locally constrained~\cite{itakura1975minimum,sakoe1978dynamic}.

% \subsection{Automatic Sound Matching}
% \label{sec:sound_matching_definition}
% Sound-matching requires a number of modular parts: the choice of synthesizer $g$, the target sounds of interest, the representation function $\phi$, and the heuristic for finding the optimal $\theta$. To give a formal definition of sound-matching, we expand the definition given by recent works~\cite{vahidi2023mesostructures,han2023perceptual}. The major components are as follows: 
% \begin{itemize}
%     \item $g(\theta)$: Parametric audio synthesizer $g$ which takes on parameters $\theta$ 
%     \item $x$: The output of $g$, given a set of parameters $\theta$ or $g(\theta) = x$ 
%     \item $t$: The target sound which we want to replicate or imitate. 
%     \item $\phi$: Representation function or feature extractor. $\phi$ is applied to $x$ and $t$ to facilitate their comparison by the loss function.
%     \item $L$: Loss or error function. $L(\theta,t)$ is a measure of distance between $x$ and $t$. Often proportional to the subtraction of their representations, or $ \phi(x) - \phi(t)$
% \end{itemize}

% \highlight{It could be argued that the implementation of these components is not an optimization problem with a correct answer, but a \textit{creative} endeavor depending on the artist's needs}. Consider the hypothetical case where two sound-designers are searching for an 8-bit~\cite{collins2007loop} version of a high quality, organic snare drum: in such a case, there is no optimal solution, and the space of possible answers is extremely large. Here, it is possible (if not highly likely) that the two artists would select different samples given a number of possible samples to choose from.

% \subsection{Past Works}
% Given that sound-matching is ultimately an imitation of a creative endeavor, it might be no surprise that it can be categorized into different subtypes.

@@

Recent years have seen more works in sound-matching using machine learning techniques. In 2018, Yee-King \textit{et al.} rendered 60,000 audio-parameter pairs from the \textit{Dexed} \gls{VST} synthesizer~\footnote{https://asb2m10.github.io/dexed/}, and showed that NNs trained on this dataset can outperform GA and hill-climber (\gls{HC})~\cite{hoffmann2000heuristic} methods in rendering speed, with slight improvements in MFCC error (used as an objective performance test). The speed improvement appears trivial, considering the iterative nature of GAs when compared to offline training of supervised models. For GAs and HC optimizer, MFCCs were used as a measure of performance as well as a loss function. For training the networks, P-Loss was used, since differentiable MFCCs were not possible in their pipeline. Importantly, informal hearing tests revealed that the performance of even the best NN model was unsatisfactory, possibly due to the complex nature of the synthesizer, which features 155 parameters.

@@

@@

>>>>

\documentclass[lettersize,journal]{IEEEtran}
\usepackage{amsmath,amsfonts}
\DeclareMathOperator*{\argmin}{arg\,min}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{array}
\usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
\usepackage{textcomp}
\usepackage{stfloats}
\usepackage{url}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{csquotes}
\usepackage{makecell}
% my added packages and commands =======================
\usepackage[numbers]{natbib} %removed cite package

@@

@@

Recent years have seen more works in sound-matching using supervised machine learning techniques. In 2018, Yee-King \textit{et al.} rendered 60,000 audio-parameter pairs from the \textit{Dexed} \gls{VST} synthesizer~\footnote{https://asb2m10.github.io/dexed/}, and showed that NNs trained on this dataset can outperform GA and hill-climber (\gls{HC})~\cite{hoffmann2000heuristic} methods in rendering speed, with slight improvements in MFCC error (used as an objective performance test). The speed improvement appears trivial, considering the iterative nature of GAs when compared to offline training of supervised models. For GAs and HC optimizer, MFCCs were used as a measure of performance as well as a loss function. For training the networks, P-Loss was used, since differentiable MFCCs were not possible in their pipeline. Importantly, informal hearing tests revealed that the performance of even the best NN model was unsatisfactory, possibly due to the complex nature of the synthesizer, which features 155 parameters.

@@

A neural approximation for another part of the supervised sound-matching chain---this time the synthesizer---was proposed by Barkan \textit{et al.}~\cite{barkan2023inversynthII}. As they noted, without a differentiable synthesizer, ``model-based'' (or what we call supervised) approaches cannot directly compare $x_o$ to $x_t$, often opting for P-Loss, which may not correctly map the parameters of sound to the output audio~\cite{esling2019flow,han2023perceptual,masuda2023improving}. Given a synthesizer and a large dataset of sounds and corresponding parameters, a model which maps sounds to parameters, and another model which approximates the synthesizer can be trained, allowing the implementation of a loss function which combines P-Loss with STFT differences. This ``Inversynth II'' (IS2) approach yielded significant improvements to previous works which did not use the STFT approximations for loss~\cite{esling2019flow,barkan2019inversynth}.

In addition, Barkan \textit{et al.} attemped to  improve the IS2 model with Inference-Time
Finetuning (ITF). For ITF, the synthesizer approximation is frozen and the encoder is iteratively finetuned for the sample sample using gradient descent in order to minimize P-Loss, stopping at convergence or when maximum number of iterations has reached~\cite{barkan2023inversynthII}. Barkan \textit{et al.} report that perhaps due to the discrepancy between the true and approximated synthesizer or overfitting, the application of ITF does not improve IS2 in spectral and manual performance metrics.

@@

% masuda and esling both noted out-of-domain drop in performance
% yee king noted performance drop in manual surveys 
%p-loss bad : masuda-improving, han2023, esling-flowsynth
